---
title: "GIS Analysis"
author: "Nicholas van Tol"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo  # Clean, modern theme from Bootstrap
    highlight: zenburn  # Darker syntax highlighting for 
    toc: true  # Include Table of Contents
    toc_depth: 2  # Include section and subsection levels
    code_folding: hide  # Initially hide code
---

## 1. Load Packages

Most of the packages used here are self-explanatory. However, the `htmltools` and `htmlwidgets` packages were used to edit aspects of the interactive map, which you will see later in the code.

```{r echo=TRUE, warning=FALSE}
# Batch Load necessary packages
packs <- c("patchwork", "tidyverse", "ctv", "rinat", "sf", "rosm", "ggspatial", 
           "prettymapr", "leaflet", "htmltools", "mapview", "leafpop", "wesanderson",
           "htmlwidgets", "htmltools", "rmarkdown")
lapply(packs, require, character.only = TRUE)
rm(packs)
```

## 2. Read Data

Start by loading the soil and vegetation databases. Then, transform the coordinate reference system to match that of the vegetation database (Transverse Mercator Lo19 - WGS 84). Afterward, read in the iNaturalist data, limiting it to 10,000 observations within a bounding box that covers the Western Cape. Finally, load a shapefile containing the borders of the Western Cape, which will be used later to limit all layers and observations within this region.

```{r}
# Read Vegetation and Soil data
veg <- st_read("NVM2024Final/Shapefile/NVM2024Final_IEM5_12_07012025.shp")
soi <- st_read("SOTER_ZA/GIS/SOTER/ZA_SOTERv1.shp")

# Reproject Soil Data to match Vegetation CRS
soi <- st_transform(soi, st_crs(veg)) 

# Inat Data: Hydnora africana & Euphorbiaceae
HA <- get_inat_obs(taxon_name = "Hydnora africana", bounds = c(-35, 18, -30, 25), maxresults = 10000)
Eu <- get_inat_obs(taxon_name = "Euphorbiaceae", bounds = c(-35, 18, -30, 25), maxresults = 10000)

# Read in Western Cape polygon and reproject
WC <- st_read("western-cape-south-africa_1334.geojson")
WC <- st_transform(WC, st_crs(veg))
```

## 3. Filter and Set Spatial Objects

Filter out iNaturalist observations that are located in the Northern Hemisphere, are cultivated, or are not research grade. Apply this filter to the dataset. Next, convert the iNaturalist observations into spatial objects and set the CRS to match that of the vegetation data. Finally, crop the data to the Western Cape using the Western Cape polygon (WC).

```{r}
# Filter observations for accuracy, latitude, and other conditions
filter_data <- function(data) {
  data %>%
    filter(latitude < 0, !is.na(latitude), 
           captive_cultivated == "false", quality_grade == "research")
}

HA <- filter_data(HA)
Eu <- filter_data(Eu)

# Convert to spatial objects
HA <- st_as_sf(HA, coords = c("longitude", "latitude"), crs = 4326)
Eu <- st_as_sf(Eu, coords = c("longitude", "latitude"), crs = 4326)

# Crop observations to Western Cape
HA <- st_transform(HA, st_crs(veg)) %>% st_intersection(WC)
Eu <- st_transform(Eu, st_crs(veg)) %>% st_intersection(WC)

# Crop vegetation and soil data to Western Cape
veg_wc <- st_intersection(veg, WC)
soi <- st_make_valid(soi)  # Fix invalid geometries
soi_wc <- st_intersection(soi, WC)
```

## 4. Display Meta Data and Live HTML

Create maps for the iNaturalist data using `mapview` and popups using `leafpop`. For each observation, create a URL link that opens the corresponding iNaturalist page. These popups will display the scientific name, user login, and the URL linking to the observation.

```{r}
# Mapview for Hydnora africana and Euphorbiaceae with popups
lpcHA <- HA %>%
  mutate(click_url = paste("<a href='", HA$url, "' target='_blank'>Link to iNat observation</a>", sep = ""))

HAm <- mapview(HA, col.regions = "purple", legend.title = "Hydnora", popup = popupTable(lpcHA, zcol = c("scientific_name", "user_login", "click_url")), layer.name = "Hydnora africana")

lpcEu <- Eu %>%
  mutate(click_url = paste("<a href='", Eu$url, "' target='_blank'>Link to iNat observation</a>", sep = ""))

Eum <- mapview(Eu, col.regions = "yellow", legend.title = "Euphorbiaceae", popup = popupTable(lpcEu, zcol = c("scientific_name", "user_login", "click_url")), layer.name = "Euphorbiaceae")

# Generate random colors for Vegetation Types
set.seed(123)
random_colors <- sample(colors(), length(unique(veg_wc$T_Name)))

# Mapview for vegetation types
veg_typ_wc <- mapview(veg_wc, zcol = "T_Name", color = random_colors, legend = FALSE, layer.name = "Vegetation Type")

# Generate random colors for Soil Types
set.seed(3000)
random_colors_soi <- sample(colors(), length(unique(soi$SOIL)))

# Mapview for soil types
soi_typ_wc <- mapview(soi_wc, zcol = "SOIL", color = random_colors_soi, legend = FALSE, layer.name = "Soil Type")
```

## 5. Buffering to Show Eu within 10km and 1km of HA

Create buffers of 10km and 1km around each Hydnora africana point. These buffers can be toggled on and off on the map. Apply filters to show only Euphorbiaceae points within these buffer regions, with different colors for each.

```{r}
# Create a 10 km and 1 km buffer around each HA point
HAbuffer_10k <- 10000  # in meters (10 km)
HAbuffer_1k <- 1000   # in meters (1 km)

# Create a 1 km and 10 km buffer around each HA point
HA_buffer10 <- st_buffer(HA, dist = HAbuffer_10k)
HA_buffer1 <- st_buffer(HA, dist = HAbuffer_1k)

# Check which Eu points fall within the 10 km and 1 km buffer of any HA points
Eu_within_HA_buffer10 <- st_intersects(Eu, HA_buffer10)
Eu_within_HA_buffer1 <- st_intersects(Eu, HA_buffer1)

# Convert the list to a logical vector where TRUE means the point is inside the buffer
Eu_within_buffer_10k <- lengths(Eu_within_HA_buffer10) > 0
Eu_within_buffer_1k <- lengths(Eu_within_HA_buffer1) > 0

# Filter Eu points that fall within the 10 km and 1 km buffer of any HA points
Eu_filtered10k <- Eu[Eu_within_buffer_10k, ]
Eu_filtered1k <- Eu[Eu_within_buffer_1k, ]

# Optionally, visualize the filtered Eu points within the buffer
HA.10k <- mapview(HA_buffer10, col.regions = "blue", alpha = 0.3, layer.name = "Hydnora africana 10km buffer")
HA.1k <- mapview(HA_buffer1, col.regions = "green", alpha = 0.3, layer.name = "Hydnora africana 1km buffer")

Eu10000HA <- mapview(Eu_filtered10k, col.regions = "red", layer.name = "Euphorbiaceae within 10km buffer")
Eu1000HA <- mapview(Eu_filtered1k, col.regions = "cyan", layer.name = "Euphorbiaceae within 1km buffer")
```

## 6. Combine All Layers

Combine all the separate maps into one large map, and then turn it into a `leaflet` object so it can be further edited using the `leaflet` package

```{r}
combined_map <- HAm + Eum + veg_typ_wc + soi_typ_wc + Eu1000HA + Eu10000HA + HA.10k + HA.1k

# for use in leaflet - add @map
comb.edit <- combined_map@map
```

## 7. Leaflet HTML Options

Use `htmltools` to remove specific CSS code that creates default zoom tabs within the maps, making the map look cleaner.

```{r}
comb.edit <- comb.edit %>%
  addControl(
    html = "<style>
              .leaflet-bar-part[title='Zoom to Hydnora africana 1km buffer'],
              .leaflet-bar-part[title='Zoom to Hydnora africana 10km buffer'],
              .leaflet-bar-part[title='Zoom to Euphorbiaceae within 10km buffer'],
              .leaflet-bar-part[title='Zoom to Euphorbiaceae within 1km buffer'],
              .leaflet-bar-part[title='Zoom to Soil Type'],
              .leaflet-bar-part[title='Zoom to Vegetation Type'],
              .leaflet-bar-part[title='Zoom to Hydnora africana'],
              .leaflet-bar-part[title='Zoom to Euphorbiaceae'] {
                display: none !important;
              }
              .state-Zoom.to.Hydnora.africana.1km.buffer.Zoom.buffer-active,
              .state-Zoom.to.Hydnora.africana.10km.buffer.Zoom.buffer-active,
              .state-Zoom.to.Euphorbiaceae.within.10km.buffer.Zoom.buffer-active,
              .state-Zoom.to.Euphorbiaceae.within.1km.buffer.Zoom.buffer-active,
              .state-Zoom.to.Soil.Type.Zoom.buffer-active,
              .state-Zoom.to.Vegetation.Type.Zoom.buffer-active,
              .state-Zoom.to.Hydnora.africana.Zoom.buffer-active,
              .state-Zoom.to.Euphorbiaceae.Zoom.buffer-active {
                display: none !important;
              }
            </style>", 
    position = "topleft", 
    className = "custom-css"
  )
```

Print the finished map with all layers. You can toggle these layers on and off using the block in the top left of the map under the +/- interface. You can also switch between different base maps.

```{r fig.width=10, fig.height=8}
comb.edit
```

## 8. Summarizing Points by Vegetation Type

Create summaries of all the points by the vegetation type they fall into. Essentially, it counts how many points fall into each vegetation type and provides an output.

```{r}
# Perform spatial join for Hydnora africana and Euphorbiaceae with vegetation
HA_points_with_vegetation <- st_join(HA, veg_wc)
Eu_points_with_vegetation <- st_join(Eu, veg_wc)

# Summarize by vegetation type
HA_summary_by_vegetation <- HA_points_with_vegetation %>%
  group_by(T_Name) %>%
  summarise(count = n(), 
            avg_x = mean(st_coordinates(HA_points_with_vegetation)[, 1]), 
            avg_y = mean(st_coordinates(HA_points_with_vegetation)[, 2]))

Eu_summary_by_vegetation <- Eu_points_with_vegetation %>%
  group_by(T_Name) %>%
  summarise(count = n(), 
            avg_x = mean(st_coordinates(Eu_points_with_vegetation)[, 1]), 
            avg_y = mean(st_coordinates(Eu_points_with_vegetation)[, 2]))

print(HA_summary_by_vegetation)

print(Eu_summary_by_vegetation)
```

## 9. Summarizing Points by Soil Type

Create summaries of all the points by the soil type they fall into. Essentially, it counts how many points fall into each soil type and provides an output.

```{r}
HA_points_with_soil <- st_join(HA, soi_wc)
Eu_points_with_soil <- st_join(Eu, soi_wc)

HA_summary_by_soil <- HA_points_with_soil %>%
  group_by(SOIL) %>%
  summarise(count = n(), 
            avg_x = mean(st_coordinates(HA_points_with_soil)[, 1]), 
            avg_y = mean(st_coordinates(HA_points_with_soil)[, 2]))

Eu_summary_by_soil <- Eu_points_with_soil %>%
  group_by(SOIL) %>%
  summarise(count = n(), 
            avg_x = mean(st_coordinates(Eu_points_with_soil)[, 1]), 
            avg_y = mean(st_coordinates(Eu_points_with_soil)[, 2]))

print(HA_summary_by_soil)

print(Eu_summary_by_soil)
```

## 10. Integrated Function

This integrated function allows the user to specify their own inputs from iNaturalist and then automatically generates a map with the observation buffers (1km and 10km), and summary tables for how many points from each taxon fall into the different vegetation and soil classes.

It simplifies and streamlines the entire process, which is detailed above (steps 1-9). Now all you need to do is specify the initial input.

```{r}
  gis_analysis <- function(taxon_names) {
    # Load necessary packages
    packs <- c("patchwork","tidyverse","ctv","rinat","sf","rosm","ggspatial","prettymapr", 
               "leaflet", "htmltools", "mapview", "leafpop", "wesanderson","htmlwidgets","htmltools","rmarkdown")
    lapply(packs, require, character.only = TRUE)
    rm(packs)
    
    # Read Vegetation and Soil data
    veg <- st_read("NVM2024Final/Shapefile/NVM2024Final_IEM5_12_07012025.shp")
    soi <- st_read("SOTER_ZA/GIS/SOTER/ZA_SOTERv1.shp")
    soi <- st_transform(soi, st_crs(veg))
    
    # Read in Western Cape polygon and reproject
    WC <- st_read("western-cape-south-africa_1334.geojson")
    WC <- st_transform(WC, st_crs(veg))
    
    # Assign colors for each taxon
    taxon_colors <- setNames(sample(colors(), length(taxon_names)), taxon_names)
    
    # Function to get and process iNaturalist data
    get_processed_inat_data <- function(taxon_name) {
      obs_data <- get_inat_obs(taxon_name = taxon_name, bounds = c(-35, 18, -30, 25), maxresults = 10000) %>%
        filter(latitude < 0, !is.na(latitude), 
               captive_cultivated == "false", quality_grade == "research") %>%
        st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
        st_transform(st_crs(veg)) %>%
        st_intersection(WC)
      return(obs_data)
    }
    
    # Get observations for each taxon
    obs_list <- lapply(taxon_names, get_processed_inat_data)
    names(obs_list) <- taxon_names
    
    # Crop vegetation and soil data to Western Cape
    veg_wc <- st_intersection(veg, WC)
    soi <- st_make_valid(soi)
    soi_wc <- st_intersection(soi, WC)
    
    # Generate mapview with popups, unique colors, and buffers
    map_layers <- list()
    buffer_1km_layers <- list()
    buffer_10km_layers <- list()
    
    for (taxon in taxon_names) {
      obs_data <- obs_list[[taxon]] %>%
        mutate(click_url = paste("<a href='", url, "' target='_blank'>Link to iNat observation</a>", sep = ""))
      
      map_layers[[taxon]] <- mapview(obs_data, col.regions = taxon_colors[[taxon]], 
                                     legend.title = taxon,  
                                     popup = popupTable(obs_data, zcol = c("scientific_name","user_login", "click_url")),
                                     layer.name = taxon)
      
      buffer_1km <- st_buffer(obs_data, dist = 1000)
      buffer_10km <- st_buffer(obs_data, dist = 10000)
      
      buffer_1km_layers[[taxon]] <- mapview(buffer_1km, col.regions = taxon_colors[[taxon]], alpha = 0.3, layer.name = paste(taxon, "1km Buffer"))
      buffer_10km_layers[[taxon]] <- mapview(buffer_10km, col.regions = taxon_colors[[taxon]], alpha = 0.2, layer.name = paste(taxon, "10km Buffer"))
    }
    
    # Random colors for Vegetation and Soil Types
    set.seed(123)
    veg_typ_wc <- mapview(veg_wc, zcol = "T_Name", color = sample(colors(), length(unique(veg_wc$T_Name))), 
                          legend = FALSE, layer.name = "Vegetation Type")
    
    set.seed(3000)
    soi_typ_wc <- mapview(soi_wc, zcol = "SOIL", color = sample(colors(), length(unique(soi$SOIL))), 
                          legend = FALSE, layer.name = "Soil Type")
    
    # Combine maps
    combined_map <- Reduce(`+`, map_layers) + Reduce(`+`, buffer_1km_layers) + Reduce(`+`, buffer_10km_layers) + veg_typ_wc + soi_typ_wc
    comb_edit <- combined_map@map
    
    # Leaflet customization
    comb_edit <- comb_edit %>%
      setView(lng = -31, lat = 21, zoom = 6) %>%
      setMaxBounds(lng1 = 17, lat1 = -35, lng2 = 25, lat2 = -30) %>% 
      addTiles(group = "OSM") %>%
      addControl(
        html = "<style>
                .leaflet-bar-part[title*='Zoom'] { display: none !important; }
                .state-Zoom.buffer-active { display: none !important; }
              </style>", 
        position = "topleft", 
        className = "custom-css"
      )
    
    # Summarizing Points by Vegetation Type
    summary_by_veg <- list()
    for (taxon in taxon_names) {
      points_with_veg <- st_join(obs_list[[taxon]], veg_wc)
      summary_by_veg[[taxon]] <- points_with_veg %>%
        group_by(T_Name) %>%
        summarise(count = n(), 
                  avg_x = mean(st_coordinates(points_with_veg)[, 1]), 
                  avg_y = mean(st_coordinates(points_with_veg)[, 2]))
    }
    
    # Summarizing Points by Soil Type
    summary_by_soil <- list()
    for (taxon in taxon_names) {
      points_with_soil <- st_join(obs_list[[taxon]], soi_wc)
      summary_by_soil[[taxon]] <- points_with_soil %>%
        group_by(SOIL) %>%
        summarise(count = n(), 
                  avg_x = mean(st_coordinates(points_with_soil)[, 1]), 
                  avg_y = mean(st_coordinates(points_with_soil)[, 2]))
    }
    
    # Save map as HTML
    saveWidget(comb_edit, "map_output.html")
    
    # Return results
    return(list(map = comb_edit, vegetation_summary = summary_by_veg, soil_summary = summary_by_soil))
  }

# Example function call
gis_analysis(c("Protea cynaroides", "Hydnora africana", "Aloe ferox"))
```
